---
title: "Stock Price Visualization using R"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---


```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.width = 7, 
                      fig.height = 5,
                      fig.align = 'center')
```

Here are some simple and effective charts to visualize stock prices. 


Load libraries:
```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(tidyquant)) install.packages("tidyquant", repos = "http://cran.us.r-project.org")
```


## Data:

Using Google stock prices from tidyquant - FANG which has historical daily prices

```{r}
dat <- tq_get("GOOG", get = "stock.prices")
```


## Line Chart:
- Using `ggplot` function with `geom_line` from `ggplot2` package to plot the line chart

```{r}
dat %>%
  ggplot(aes(x = date, y = close)) +
  geom_line() +
  labs(title = "Google", 
       subtitle = "Closing Prices - Line Chart", 
       x = "Year",
       y = "Closing Prices") 
```


## Candlestick Chart:
- Using `geom_candlestick` from `tidyquant` to plot Candlestick chart
- Candlestock chart needs Open,High,Low and Close prices
- Line colors are set using `colour-up` & `colour_down`
- Fill colors are set using `fill_up` & `fill_down`


```{r}
dat$year <- year(dat$date)

dat %>%
  filter(year==2019) %>%
  ggplot(aes(x = date, y = close)) +
   geom_candlestick(aes(open = open, high = high, low = low, close = close),
                        colour_up = "green4", colour_down = "red", 
                        fill_up  = "green4", fill_down  = "red") +
  labs(title = "Google", 
       subtitle = "Closing Prices - Candlestick Chart",
       x = "Date", 
       y = "Closing Price")
```

### Zooming in on the Candlestick Chart
-  using `coord_x_date` to zoom in by specifying the start date and end date for x-axis and price range for the y-axis  

```{r}
enddate <- max(dat$date)
startdate <- enddate - weeks(4)

dat %>%
  ggplot(aes(x = date, y = close)) +
     geom_candlestick(aes(open = open, high = high, low = low, close = close),
                        colour_up = "green4", colour_down = "red", 
                        fill_up  = "green4", fill_down  = "red") +
    labs(title = "Google", 
         subtitle = "Zoomed in Candlestick Chart",
          x = "Date",
         y = "Closing Price") + 
    coord_x_date(xlim = c(startdate, enddate),
                 ylim = c(1350, 1500)) 
```


## Calendar Heatmap:
- a heatmap of daily prices shown in a calendar 

Adding required variables for calendar heatmap

```{r}
dat$year <- year(dat$date)
dat$month <- month(dat$date)
dat$weekday = as.POSIXlt(dat$date)$wday 
dat$weekofmonth= (1 + as.integer(format(dat$date, "%d")) %/% 7)

# to order by month names
dat$month.f<-factor(dat$month,levels=as.character(1:12),
                    labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"),
                    ordered=TRUE)  

# to order by week day names
dat$weekday.f<-factor(dat$weekday,levels=rev(1:7),
                      labels=rev(c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")),
                      ordered=TRUE)  

```


- using `facet_grid` to create year, month grids 
- using `scale_fill_gradient` to show low to high prices in specified color gradient

```{r, fig.height=10}
dat %>% ggplot(aes(weekofmonth, weekday.f, fill = close)) + 
  geom_tile(colour = "white") + 
  facet_grid(year~month.f) + 
  scale_fill_gradient(low="red", high="green4") +
  labs(title = "Google", 
       subtitle = "Closing Prices Heatmap",
       x = "Week of Month",
       y = "Week Day",
       fill = "Close")

```


[R-Markdown](https://github.com/madankundapur/DataAnalytics/blob/master/StockPriceVisualization.Rmd)
